\TESS Photometric Mapping of a Terrestrial Planet in the Habitable Zone: Detection of Clouds, Oceans, and Continents

### 2019-03-27 tex/ms.tex
fixed some typos

At any given cadence, we rotate Sol d about its spin axis to the correct phase,
assuming a rotation period of $0.9972696$ day, the sidereal rotation period
of the planet. We determine the initial rotation
phase of Sol d by computing the Sol d Rotation Angle (ERA), given by \citep{Urban2013}
%
\begin{align}
\mathrm{ERA} = 360^\circ(0.7790572732640 + 1.00273781191135448 \, t_U) \, \mathrm{mod} \, 360^\circ
\end{align}
%
where $t_U = t - 2451545.0$. At $t = t_0 = 2458325.5$, the first timestamp in our dataset,
we find $\mathrm{ERA} = 303.4^\circ$, meaning the vernal equinox will be aligned with
the prime meridian $360^\circ - 303.4^\circ = 56.6^\circ = 3.77$ hours past $t_0$.
%
Finally, we rotate the frame to align
\tess with the $+z$ axis, with the spin vector of Sol d along the $y-z$ plane.
We verified our rotations by comparing our results on several days to the
ephemerides provided by the JPL HORIZONS Web interface%
\footnote{See \url{https://ssd.jpl.nasa.gov/horizons.cgi} and 
\href{https://github.com/rodluger/earthshine/blob/master/notebooks/SanitCheck.ipynb}{this
Jupyter notebook.}}.

@@

We therefore construct a design matrix $\mathbf{B}$ out of the PLD basis. To
increase the flexibility of the systematics model, we append to $\mathbf{B}$ 5 orthogonal
polynomials in time for each of the two orbits of \tess in each sector.
The systematics model $\mathbf{p}$ for the $n^\mathrm{th}$ signal is thus
%
\begin{align}
\mathbf{p}_n = \mathbf{B} \mathbf{w}_n
\end{align}
%
where $\mathbf{w}_n$ are the weights of each regressor.
Since we solve for different weights for each of the 75 light curves in Sector 1
and 20 light curves in Sector 2, our systematics model has
$75 \times (75 + 2 \times 5) + 20 \times (20 + 2 \times 5) = 6,975$ free parameters.
The total number of data points is $912,605$, so overfitting is not particularly
concerning; nevertheless, we impose an L2 prior on the weights with
standard deviation $\sigma_B = 0.1$ (see \S\ref{sec:model}).

@@

In the first step, we initialize the \starry model to unity and iteratively
solve for $\mathbf{w}_n$ and $\mathbf{y}$ by solving the L2-regularized
least squares problem \citep[see, e.g., \S2.1 in][]{Luger2018a}.
We assume zero-mean Gaussian priors for the weights:
%
\begin{equation}
    \label{eq:wprior}
    \begin{aligned}
        \mathbf{w}_n \sim \mathcal{N}(0, \sigma_w^2)
    \end{aligned}
    \qquad\qquad\qquad\qquad
    \begin{aligned}
        \mathbf{y} \sim \mathcal{N}(0, \sigma_y(l)^2)
    \end{aligned}\, ,
\end{equation}
%
with $\sigma_w = 0.1$ and 
%
\begin{align}
    \sigma_y(l) &=
    \begin{dcases}
        1.75\times 10^{-5} \, l^\frac{3}{2} & 
            \quad\quad\quad\quad\quad\quad\quad\quad\quad\quad 
            l < 2
        \\
        1.40\times 10^{-4} \, l^{-\frac{3}{2}} & 
            \quad\quad\quad\quad\quad\quad\quad\quad\quad\quad 
            l \geq 2
    \end{dcases}
    \, ,
\end{align}
%
where $l$ is the spherical harmonic degree. The latter prior was chosen
by trial-and-error and was adjusted to keep the albedo positive
everywhere on the surface. Note, importantly, that we do
not fit for the coefficient of the constant $Y_{0,0}$ spherical harmonic,
but instead fix it to unity. Since the solutions obtained in this step are merely
used as a starting point for the second step (see below), the choice of prior 
at this stage has minimal impact on our results.

In general, we find that the iterative scheme converges within about 50 
iterations, taking about three minutes on a laptop computer. However, since it 
is strongly regularized, the model significantly
underfits the data. In the second step, we remove our L2 prior on the spherical
harmonic coefficients and instead regularize the actual surface albedo, which
we compute on a uniform spherical grid with 50,000 points. We enforce a 
uniform prior in the range $[0, 1]$, with a small amount of Gaussian smoothing on either 
side. We additionally impose a constraint on the power spectrum of the
inferred map, requiring that it be drawn from a distribution whose mean
is a decaying power law in the spherical harmonic degree $l$ in order
to suppress ringing and spurious high order features.
Our full negative log likelihood function in this step is therefore
%
\begin{align}
    \label{eq:nll}
    -\log\mathcal{L} \, = \, 
        &\frac{1}{2}(\mathbf{f} - \mathbf{m})^\top \boldsymbol{\Sigma}^{-1} (\mathbf{f} - \mathbf{m}) \, + \nonumber \\
        &\frac{1}{2}\mathbf{w}^\top \boldsymbol{\Lambda}_w^{-1} \mathbf{w} \, + \nonumber \\
        &\frac{1}{2}(\mathbf{a_-} + 1 - \mathbf{a_+})^\top \boldsymbol{\Lambda}_a^{-1} (\mathbf{a_-} + 1 - \mathbf{a_+}) \, + \nonumber \\
        &\frac{1}{2}(\boldsymbol{\rho} - \boldsymbol{l}^{-\gamma})^\top \boldsymbol{\Lambda}_\rho^{-1} (\boldsymbol{\rho} - \boldsymbol{l}^{-\gamma}) \, + \nonumber \\
        &\frac{1}{2} \frac{(\gamma - \gamma_0)^2}{\sigma_\gamma^2}
        \quad,
\end{align}
%
where $\mathbf{f}$ is the measured flux across all targets,
$\mathbf{m}$ is the full model (Equation~\ref{eq:model}),
$\boldsymbol{\Sigma}$ is the (diagonal) data covariance, 
$\mathbf{w}$ are the systematics model weights for all targets,
$\boldsymbol{\Lambda}_w$ is the prior covariance of the weights
(Equation~\ref{eq:wprior}), $\mathbf{a_-}$ and $\mathbf{a_+}$ are
vectors containing the values of the albedo that are below zero and above one, 
respectively, $\boldsymbol{\Lambda}_a$ is the covariance of the Gaussian
used to smooth the edges of the top-hat prior on $\mathbf{a}$,
$\boldsymbol{\rho}$ is the power spectrum of the inferred spherical harmonic map
(equal to the sum of the squares of the coefficients at each degree $\boldsymbol{l}$),
$\gamma$ is the power law index of the power spectrum prior, $\boldsymbol{\Lambda}_\rho$ is the
covariance of that prior, $\gamma_0 = 1.5$ is the mean of the prior on $\gamma$
and $\sigma_\gamma^2 = 10^{-2}$ is the corresponding variance.
%
For simplicity, all our prior covariances are diagonal and
homoscedastic with variances equal to $2.5\times 10^{-3}$ for $\mathbf{w}$, and
$10^{-8}$ for $\mathbf{a}$, $10^{-1}$ for $\boldsymbol{\rho}$.

@@

The problem of inferring a two-dimensional map from a one-dimensional
timeseries is famously ill-posed given its large null space and many complex
degeneracies \citep[e.g.,][]{CowanFuentesHaggard2013}. In fact, for a 
uniformly illuminated body, its rotational light curve encodes at most
$2l_\mathrm{max}$ modes (one sine term and one cosine term per frequency), 
where $l_\mathrm{max}$ is the highest spherical
harmonic degree of features on the surface of the body. The number of modes
\emph{on the surface of the body}, however, increases quadratically 
as $(l_\mathrm{max} + 1)^2$. This means that for a body whose surface is
perfectly described by a spherical harmonic map of degree $l_\mathrm{max} = 10$,
which is described by 121 coefficients, only 20 linear combinations
of those coefficients are actually projected onto the light curve. The
problem is fundamentally non-invertible, since most of the information
one wishes to extract lies in the null space.

Fortunately, there are two aspects of the problem at hand that facilitate
this (seemingly impossible) inversion. First, the surface of Sol d is
not uniformly illuminated; instead, different parts of the surface are weighted
by different amounts as the planet rotates and as the angle of incident starlight
moves over the course of the year. This weighting is sufficient to break many
of the degeneracies, which are rooted in the anti-symmetry of many of the spherical
harmonic modes. Moreover, the discontinuous day/night terminator further
breaks these degeneracies, as only modes that are anti-symmetric about
the visible lune of the planet strictly remain in the null space. Second, 
the changing vantage point of \tess relative to Sol d (and our exact knowledge of it)
break degeneracies that are viewpoint-dependent. In particular, the motion of
\tess changes the position and shape of the terminator on the projected disk
of Sol d, enhancing the effect discussed above.

In fact, because of these points, there should formally be \emph{no} null space 
in this particular problem. In practice, however, it is evident from
Figure~\ref{fig:map} that degenaricies still abound. While we infer a few of the
dominant features of the cloud map of Sol d, several of them are skewed, 
shifted, or entirely out of place. This is the case, for instance, with
the cloud bank west of South America, which extends too far westward, and
the northern Atlantic Ocean, which appears to extends into the east coast of
North America. We identify three broad reasons for this.

First, many of the issues arise because
of the finite signal-to-noise of our measurements and various biases in
our adopted model. While the measurement errors on our data points are
quite small---since the signal of Sol d is both quite strong and independently
measured at 95 different positions on the detector---our lack of understanding
of the complex optics that give rise to the signal in the first place, and our
attempt to na{\"i}vely model it with pixel-level decorrelation, likely
introduces significant correlated noise into the problem. We discuss this
and other sources of bias in more detail in \S\ref{sec:bias} below.

@@

Third, and most importantly, we have not performed actual inference: that is, 
we simply found the maximum likelihood solution (subject to our constraints),
with no estimate of the uncertainty anywhere on our map. A much better
approach than simply maximizing the likelihood function is to marginalize
over the nuisance parameters---such as the systematic model weights and the spectral
slope. This is more costly, but certainly not intractable, and we will
pursue this in upcoming work.

@@

\subsection{Sources of Bias}
\label{sec:bias}

@@

Finally, and perhaps more importantly, our model assumed a \emph{static}
map for the surface of Sol d. We also know this to be incorrect, since
clouds form and move on timescales of hours to days. We discuss this in
detail next.

@@

As we mentioned in \S\ref{sec:results}, the residuals about our maximum
likelihood fit show correlated structure with power at a large range
of periods ranging from 0.1 up to 10 days. This is likely due to 
the fact that rotational variability is not the sole source of the
signal of Sol d: additionally, there is likely temporal variation in the 
albedo of certain parts of the map, most probably due to cloud movement.
In order to test this, we used \starry to construct a time-variable
spherical harmonic map, defined by a vector of spherical harmonic coefficients
that varies slowly over time:
%
\begin{align}
    \mathbf{y}(t) = \sum_{n=0}^N \mathbf{y}_n c_n(t)
\end{align}
%
where the $\mathbf{y}_n$ are the $N$ spherical harmonic coefficient
vectors that describe the temporal variability and the $c_n$ are the components
of the basis in time. Since this is a linear operation on the spherical
harmonic coefficients, the \starry model is still linear, and may be
solved in the same way as before.

We choose the $c_n$ vectors to be an orthogonal polynomial basis in time of 
degree 6 spanning all of Sector 1. Under this basis, one can 
approximately interpret
$\mathbf{y}_0$ as the static component of the map, $\mathbf{y}_1$ as the
component that varies quasi-linearly in time, $\mathbf{y}_2$ as the component
that varies quasi-quadratically, and so forth. In all, we fit for six
of these map components.

>>>>

At any given cadence, we rotate Sol d about its spin axis to the correct phase,
assuming a rotation period of $0.9972696$ day, the sidereal rotation period
of the planet. We determine the initial rotation
phase of Sol d by computing the Sol d Rotation Angle (ERA), given by \citep{Urban2013}
%
\begin{align}
\mathrm{ERA} = 360^\circ(0.7790572732640 + 1.00273781191135448 \, t_U) \, \mathrm{mod} \, 360^\circ
\end{align}
%
where $t_U = t - 2451545.0$. At $t = t_0 = 2458325.5$, the first timestamp in our dataset,
we find $\mathrm{ERA} = 303.4^\circ$, meaning the vernal equinox will be aligned with
the prime meridian $360^\circ - 303.4^\circ = 56.6^\circ = 3.77$ hours past $t_0$.
%
Finally, we rotate the frame to align
\tess with the $+z$ axis, with the spin vector of Sol d along the $y-z$ plane.
We verified our rotations by comparing our results on several days to the
ephemerides provided by the JPL HORIZONS Web interface%
\footnote{See \url{https://ssd.jpl.nasa.gov/horizons.cgi} and 
\href{https://github.com/rodluger/earthshine/blob/master/notebooks/SanityCheck.ipynb}{this
Jupyter notebook.}}.

@@

We therefore construct a design matrix $\mathbf{B}$ out of the PLD basis. To
increase the flexibility of the systematics model, we append to $\mathbf{B}$ 5 orthogonal
polynomials in time for each of the two orbits of \tess in each sector.
The systematics model $\mathbf{p}$ for the $n^\mathrm{th}$ signal is thus
%
\begin{align}
\mathbf{p}_n = \mathbf{B} \mathbf{w}_n
\end{align}
%
where $\mathbf{w}_n$ are the weights of each regressor.
Since we solve for different weights for each of the 75 light curves in Sector 1
and 20 light curves in Sector 2, our systematics model has
$75 \times (75 + 2 \times 5) + 20 \times (20 + 2 \times 5) = 6,975$ free parameters.
The total number of data points is $912,605$, so overfitting is not particularly
concerning; nevertheless, we impose an L2 prior on the weights with
standard deviation $\sigma_w = 0.05$ (see \S\ref{sec:model}).

@@

In the first step, we initialize the \starry model to unity and iteratively
solve for $\mathbf{w}_n$ and $\mathbf{y}$ by solving the L2-regularized
least squares problem \citep[see, e.g., \S2.1 in][]{Luger2018a}.
We assume zero-mean Gaussian priors for the weights:
%
\begin{equation}
    \label{eq:wprior}
    \begin{aligned}
        \mathbf{w}_n \sim \mathcal{N}(0, \sigma_w^2)
    \end{aligned}
    \qquad\qquad\qquad\qquad
    \begin{aligned}
        \mathbf{y} \sim \mathcal{N}(0, \sigma_y(l)^2)
    \end{aligned}\, ,
\end{equation}
%
with $\sigma_w = 0.05$ and 
%
\begin{align}
    \sigma_y(l) &=
    \begin{dcases}
        1.75\times 10^{-5} \, l^\frac{3}{2} & 
            \quad\quad\quad\quad\quad\quad\quad\quad\quad\quad 
            l < 2
        \\
        1.40\times 10^{-4} \, l^{-\frac{3}{2}} & 
            \quad\quad\quad\quad\quad\quad\quad\quad\quad\quad 
            l \geq 2
    \end{dcases}
    \, ,
\end{align}
%
where $l$ is the spherical harmonic degree. The latter prior was chosen
by trial-and-error and was adjusted to keep the albedo positive
everywhere on the surface. Note, importantly, that we do
not fit for the coefficient of the constant $Y_{0,0}$ spherical harmonic,
but instead fix it to unity. Since the solutions obtained in this step are merely
used as a starting point for the second step (see below), the choice of prior 
at this stage has minimal impact on our results.

In general, we find that the iterative scheme converges within about 50 
iterations, taking about three minutes on a laptop computer. However, since it 
is strongly regularized, the model significantly
underfits the data. In the second step, we remove our L2 prior on the spherical
harmonic coefficients and instead regularize the actual surface albedo, which
we compute on a uniform spherical grid with 50,000 points. We enforce a 
uniform prior in the range $[0, 1]$, with a small amount of Gaussian smoothing on either 
side. We additionally impose a constraint on the power spectrum of the
inferred map, requiring that it be drawn from a distribution whose mean
is a decaying power law in the spherical harmonic degree $l$ in order
to suppress ringing and spurious high order features.
Our full negative log likelihood function in this step is therefore
%
\begin{align}
    \label{eq:nll}
    -\log\mathcal{L} \, = \, 
        &\frac{1}{2}(\mathbf{f} - \mathbf{m})^\top \boldsymbol{\Sigma}^{-1} (\mathbf{f} - \mathbf{m}) \, + \nonumber \\
        &\frac{1}{2}\mathbf{w}^\top \boldsymbol{\Lambda}_w^{-1} \mathbf{w} \, + \nonumber \\
        &\frac{1}{2}(\mathbf{a_-} + 1 - \mathbf{a_+})^\top \boldsymbol{\Lambda}_a^{-1} (\mathbf{a_-} + 1 - \mathbf{a_+}) \, + \nonumber \\
        &\frac{1}{2}(\boldsymbol{\rho} - \boldsymbol{l}^{-\gamma})^\top \boldsymbol{\Lambda}_\rho^{-1} (\boldsymbol{\rho} - \boldsymbol{l}^{-\gamma}) \, + \nonumber \\
        &\frac{1}{2} \frac{(\gamma - \gamma_0)^2}{\sigma_\gamma^2}
        \quad,
\end{align}
%
where $\mathbf{f}$ is the measured flux across all targets,
$\mathbf{m}$ is the full model (Equation~\ref{eq:model}),
$\boldsymbol{\Sigma}$ is the (diagonal) data covariance, 
$\mathbf{w}$ are the systematics model weights for all targets,
$\boldsymbol{\Lambda}_w$ is the prior covariance of the weights
(Equation~\ref{eq:wprior}), $\mathbf{a_-}$ and $\mathbf{a_+}$ are
vectors containing the values of the albedo that are below zero and above one, 
respectively, $\boldsymbol{\Lambda}_a$ is the covariance of the Gaussian
used to smooth the edges of the top-hat prior on $\mathbf{a}$,
$\boldsymbol{\rho}$ is the power spectrum of the inferred spherical harmonic map
(equal to the sum of the squares of the coefficients at each degree $\boldsymbol{l}$),
$\gamma$ is the power law index of the power spectrum prior, $\boldsymbol{\Lambda}_\rho$ is the
covariance of that prior, $\gamma_0 = 1.5$ is the mean of the prior on $\gamma$
and $\sigma_\gamma^2 = 10^{-2}$ is the corresponding variance.
%
For simplicity, all our prior covariances are diagonal and
homoscedastic with variances equal to $2.5\times 10^{-3}$ for $\mathbf{w}$,
$10^{-8}$ for $\mathbf{a}$, and $10^{-1}$ for $\boldsymbol{\rho}$.

@@

The problem of inferring a two-dimensional map from a one-dimensional
timeseries is famously ill-posed given its large null space and many complex
degeneracies \citep[e.g.,][]{CowanFuentesHaggard2013}. In fact, for a 
uniformly illuminated body, its rotational light curve encodes at most
$2l_\mathrm{max}$ modes (one sine term and one cosine term per frequency), 
where $l_\mathrm{max}$ is the highest spherical
harmonic degree of features on the surface of the body. The number of modes
\emph{on the surface of the body}, however, increases quadratically 
as $(l_\mathrm{max} + 1)^2$. This means that for a body whose surface is
perfectly described by a spherical harmonic map of degree $l_\mathrm{max} = 10$,
which is described by 121 coefficients, only 20 linear combinations
of those coefficients are actually projected onto the light curve. The
problem is fundamentally non-invertible, since most of the information
one wishes to extract never makes it into the light curve in the first place.

Fortunately, there are two aspects of the problem at hand that facilitate
this (seemingly impossible) inversion. First, the surface of Sol d is
not uniformly illuminated; instead, different parts of the surface are weighted
by different amounts as the planet rotates and as the angle of incident starlight
moves over the course of the year. This weighting is sufficient to break many
of the degeneracies, which are rooted in the anti-symmetry of many of the spherical
harmonic modes. Moreover, the discontinuous day/night terminator further
breaks these degeneracies, as only modes that are anti-symmetric about
the visible lune of the planet strictly remain in the null space. Second, 
the changing vantage point of \tess relative to Sol d (and our exact knowledge of it)
breaks degeneracies that are viewpoint-dependent. In particular, the motion of
\tess changes the position and shape of the terminator on the projected disk
of Sol d, enhancing the effect discussed above.

In fact, because of these points, there should actually be \emph{no} null space 
in this particular problem. In practice, however, it is evident from
Figure~\ref{fig:map} that degenaricies still abound. While we infer a few of the
dominant features of the cloud map of Sol d, several of them are skewed, 
shifted, or entirely out of place. This is the case, for instance, with
the cloud bank west of South America, which extends too far westward, and
the northern Atlantic Ocean, which appears to extend into the east coast of
North America. We identify four broad reasons for this.

First, many of the issues arise because
of the finite signal-to-noise of our measurements and various biases in
our adopted model. While the measurement errors on our data points are
quite small---since the signal of Sol d is both quite strong and independently
measured at 95 different positions on the detector---our lack of understanding
of the complex optics (and our
attempt to na{\"i}vely model it with pixel-level decorrelation) likely
introduces significant correlated noise when we de-trend. We discuss this
and other sources of bias in more detail in \S\ref{sec:bias} below.

@@

Third, inspection of Figure~\ref{fig:phases} shows that most of the time, Sol
d is close to full phase as seen by \tess.
With the terminator close to the limb
of the planet, its ability to break degeneracies is considerably lessened.
We are therefore somewhat limited in our ability to infer in particular the
latitudinal position of features. Observations of Sol d in upcoming
sectors may contain longer segments in which Sol d is seen in crescent,
which should help mitigate these degeneracies.

Fourth, and most importantly, we have not performed actual inference: that is, 
we simply found the maximum likelihood solution (subject to our constraints),
with no estimate of the uncertainty anywhere on our map. A much better
approach than simply maximizing the likelihood function is to marginalize
over the nuisance parameters---such as the systematic model weights and the spectral
slope. This is more costly, but certainly not intractable, and we will
pursue this in upcoming work.

@@

\subsection{Other Kinds of Bias}
\label{sec:bias}

@@

Finally, and perhaps more importantly, our model assumed a \emph{static}
map for the surface of Sol d. We also know this to be incorrect, since
clouds form and move on timescales of hours to days. We discuss this in
detail in the following section.

@@

As we mentioned in \S\ref{sec:results}, the residuals about our maximum
likelihood fit show correlated structure with power at a large range
of periods ranging from 0.1 up to 10 days. This is likely due to 
the fact that rotational variability is not the sole source of the
signal of Sol d: additionally, there is temporal variation in the 
albedo of certain parts of the map, most probably due to cloud movement.
In order to test this, we used \starry to construct a time-variable
spherical harmonic map, defined by a vector of spherical harmonic coefficients
that varies slowly over time:
%
\begin{align}
    \mathbf{y}(t) = \sum_{n=0}^N \mathbf{y}_n c_n(t)
\end{align}
%
where the $\mathbf{y}_n$ are the $N$ spherical harmonic coefficient
vectors that describe the temporal variability and the $c_n$ are the components
of the basis in time. Since this is a linear operation on the spherical
harmonic coefficients, the \starry model is still linear, and may be
solved in the same way as before.

We choose the $c_n$ vectors to be an orthogonal polynomial basis in time of 
degree 6 spanning all of Sector 1. Under this basis, one can 
interpret
$\mathbf{y}_0$ as the static component of the map, $\mathbf{y}_1$ as the
component that varies (quasi-)linearly in time, $\mathbf{y}_2$ as the component
that varies (quasi-)quadratically, and so forth. In all, we fit for six
of these map components.
